---
title: "Quiz 4"
author: "Tushar Kataria"
output:
  pdf_document:
    highlight: tango
    number_sections: true
  html_document:
    theme: flatly
header-includes:
- \usepackage{titling}
- "\\setlength{\\droptitle}{-6em}"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Question 1

A pharmaceutical company is interested in testing a potential blood pressure lowering medication. Their first examination considers only subjects that received the medication at baseline then two weeks later. The data are as follows (SBP in mmHg)

| Subject | Baseline | Week 2 |
|---------|----------|--------|
| 1       | 140      | 132    |
| 2       | 138      | 135    |
| 3       | 150      | 151    |
| 4       | 148      | 146    |
| 5       | 135      | 130    |

Consider testing the hypothesis that there was a mean reduction in blood pressure? Give the P-value for the associated two sided T test.

(Hint, consider that the observations are paired.)

## Answer

Here, for a mean difference between the followup and baseline of $\mu_{d}$, the null hypothesis is $H_{0} : \mu_{d} = 0$ which says that there is no difference in the blood pressure while the alternative hypothesis is $H_{a} : \mu_{d} \neq 0$ (there is a difference). So, the p-value for the associated two-sided t-test for these paired observations is as follows:

```{r q1}
baseline <- c(140, 138, 150, 148, 135)
followup <- c(132, 135, 151, 146, 130)

t.test(followup, baseline, alternative = "two.sided", paired = TRUE)$p.value
```

# Question 2

A sample of 9 men yielded a sample average brain volume of 1,100cc and a standard deviation of 30cc. What is the complete set of values of $\mu_{0}$ that a test of $H_{0} : \mu = \mu_{0}$ would fail to reject the null hypothesis in a two sided 5% Students t-test?

## Answer

```{r q2}
1100 + c(-1, 1) * qt(0.975, 8) * 30/sqrt(9)
```

# Question 3

Researchers conducted a blind taste test of Coke versus Pepsi. Each of four people was asked which of two blinded drinks given in random order that they preferred. The data was such that 3 of the 4 people chose Coke. Assuming that this sample is representative, report a P-value for a test of the hypothesis that Coke is preferred to Pepsi using a one sided exact test.

## Answer

Let $p$ be the proportion of people who prefer Coke, such that the null hypothesis is $H_{0} : p = 0.5$ versus the alternative hypothesis $H_{a} : p > 0.5$.

Let $X$ be the number of out of 4 that prefer Coke and assume $X \sim Binomial(p, 0.5)$.

With this, the p-value is calculated as

$$
\begin{aligned}
  p-value = P(X \geq 3) &= \binom{4}{3} 0.5^{3} 0.5^{1} + \binom{4}{4} 0.5^{4} 0.5^{0} \\
  &= \bf{\binom{4}{3} 0.5^{4} + \binom{4}{3} 0.5^{4}}
\end{aligned}
$$

This is computed in R using the binomial distribution as shown below:

```{r q3.1}
pbinom(2, size = 4, prob = 0.5, lower.tail = FALSE)
```

Alternatively, the equation above can also be used to calculate this p-value:

```{r q3.2}
choose(4, 3) * 0.5^4 + choose(4, 4) * 0.5^4
```

# Question 4

Infection rates at a hospital above 1 infection per 100 person days at risk are believed to be too high and are used as a benchmark. A hospital that had previously been above the benchmark recently had 10 infections over the last 1,787 person days at risk. About what is the one sided P-value for the relevant test of whether the hospital is *below* the standard?

## Answer

The null and alternative hypotheses are $H_{0} : \lambda = 0.01$ versus $H_{a} : \lambda < 0.01$ with $X = 11$ and $t = 1787$, and assuming $X \sim Poisson(0.01 \times t)$, the one-sided p-value is

```{r q4}
ppois(10, lambda = 0.01 * 1787)
```

# Question 5

Suppose that 18 obese subjects were randomized, 9 each, to a new diet pill and a placebo. Subjects’ body mass indices (BMIs) were measured at a baseline and again after having received the treatment or placebo for four weeks. The average difference from follow-up to the baseline (followup - baseline) was -3 kg/m2 for the treated group and 1 kg/m2 for the placebo group. The corresponding standard deviations of the differences was 1.5 kg/m2 for the treatment group and 1.8 kg/m2 for the placebo group. Does the change in BMI appear to differ between the treated and placebo groups? Assuming normality of the underlying data and a common population variance, give a pvalue for a two sided t test.

## Answer

The null hypothesis is $H_{0} : \mu_{difference, treated} = \mu_{difference, placebo}$

```{r q5}
n_t <- n_p <- 9

# treated
mu_t <- -3
s_t <- 1.5

# placebo
mu_p <- 1
s_p <- 1.8

s_pool <- sqrt(((n_t - 1) * s_t^2 + (n_p - 1) * s_p^2) / (n_t + n_p - 2))

ts <- (mu_t - mu_p) / (s_pool * sqrt(1/n_t + 1/n_p))
2 * pt(ts, n_t + n_p - 2)
```

Therefore, the p-value for a two-sided t-test is **less than 0.01**.

# Question 6

Brain volumes for 9 men yielded a 90% confidence interval of 1,077 cc to 1,123 cc. Would you reject in a two sided 5% hypothesis test of $H_{0} : \mu = 1,078$?

## Answer

**No, you wouldn't reject.** The 95% interval would be wider than the 90% interval. Since 1,078 is in the narrower 90% interval, it would also be in the wider 95% interval. Thus, in either case it's in the interval and so you would fail to reject.

# Question 7

Researchers would like to conduct a study of 100 healthy adults to detect a four year mean brain volume loss of $.01 mm^3$. Assume that the standard deviation of four year volume loss in this population is $.04 mm^3$. About what would be the power of the study for a 5%, percent one sided test versus a null hypothesis of no volume loss?

## Answer

The hypothesis is $H_{0} : \mu_{\Delta} = 0$ versus $H_{a} : \mu_{\Delta} > 0$ where $\mu_{\Delta}$ is volume loss (defined as Baseline - Four Weeks). The test statistic is $10 \frac{\bar{X}_{\Delta}}{0.04}$ which is rejected if it is larger than $Z_{0.95} = 1.645$.

We want to calculate

$$
\begin{aligned}
  P \left( \frac{\bar{X}_{\Delta}}{\sigma_{\Delta} / 10} > 1.645 | \mu_{\Delta = 0.01} \right) &= P \left( \frac{\bar{X}_{\Delta} - 0.01}{0.004} > 1.645 - \frac{0.01}{0.004} | \mu_{\Delta = 0.01} \right) \\
  &= P(Z > -0.855) \\
  &= 0.80
\end{aligned}
$$

Or note that $\bar{X}_{\Delta}$ is $N(0.01, 0.004)$ under the alternative and we want $P(\bar{X}_{\Delta} > 1.645 \times 0.004)$ under $H_{a}$.

```{r q7.1}
pnorm(1.645 * 0.004, mean = 0.01, sd = 0.004, lower.tail = FALSE)
```

Alternatively, the power can be calculated using `power.t.test` in R for a similar result:

```{r q7.2}
power.t.test(n = 100, delta = 0.01 - 0, sd = 0.04, type = "one.sample", alt = "one.sided")$power
```

# Question 8

Researchers would like to conduct a study of $n$ healthy adults to detect a four year mean brain volume loss of $.01 mm^3$. Assume that the standard deviation of four year volume loss in this population is $.04 mm^3$. About what would be the value of n needed for 90%, percent power of type one error rate of 5%, percent one sided test versus a null hypothesis of no volume loss?

## Answer

The hypothesis is $H_{0} : \mu_{\Delta} = 0$ versus $H_{a} : \mu_{\Delta} > 0$ where $\mu_{\Delta}$ is volume loss (defined as Baseline - Four Weeks). The test statistic is $10 \frac{\bar{X}_{\Delta}}{0.04 / \sqrt{n}}$ which is rejected if it is larger than $Z_{0.95} = 1.645$.

We want to calculate

$$
\begin{aligned}
  P \left( \frac{\bar{X}_{\Delta}}{\sigma_{\Delta} / \sqrt{n}} > 1.645 | \mu_{\Delta = 0.01} \right) &= P \left( \frac{\bar{X}_{\Delta} - 0.01}{0.04 / \sqrt{n}} > 1.645 - \frac{0.01}{0.04 / \sqrt{n}} | \mu_{\Delta = 0.01} \right) \\
  &= P(Z > 1.645 - \sqrt{n} / 4) \\
  &= 0.90
\end{aligned}
$$

So we need $1.645 - \sqrt{n} / 4 = Z_{0.10} = -1.282$ and thus

$$
\begin{aligned}
  n = [4 \times (1.645 + 1.282)] ^ 2
\end{aligned}
$$

```{r q8.1}
ceiling( (4 * ( qnorm(0.95) - qnorm(0.1) )) ^ 2 )
```

Alternatively, `power.t.test` in R can be used to get the sample size:

```{r q8.2}
power.t.test(power = 0.9, delta = 0.01 - 0, sd = 0.04, type = "one.sample", alt = "one.sided")$n
```

# Question 9

As you increase the type one error rate, $\alpha$, what happens to power?

## Answer

**You will get larger power** because as you require less evidence to reject (your $\alpha$ rate goes up), you will have larger power.
